<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{vignette 1: capture first match}
-->

# Capture first match

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Capture first match in each character vector element

Consider the following vector which contains genome position strings,

```{r}
pos.vec <- c(
  "chr10:213,054,000-213,055,000",
  "chrM:111,000",
  "chr1:110-111 chr2:220-222") # two possible matches.
```

To capture the first genome position in each string, we use the following syntax.
The first argument is the subject character vector, 
and the other arguments are pasted together 
to make a capturing regular expression.
Each named argument generates a capture group; 
the R argument name is used for the column name of the result.

```{r}
(chr.dt <- nc::capture_first_vec(
  pos.vec, 
  chrom="chr.*?",
  ":",
  chromStart="[0-9,]+"))
str(chr.dt)
```

We can add type conversion functions on the same line as each named argument:

```{r}
keep.digits <- function(x)as.integer(gsub("[^0-9]", "", x))
(int.dt <- nc::capture_first_vec(
  pos.vec, 
  chrom="chr.*?",
  ":",
  chromStart="[0-9,]+", keep.digits))
str(int.dt)
```

Below we use list variables to create patterns which are re-usable,
and we use an un-named list to generate a non-capturing optional group:

```{r}
pos.pattern <- list("[0-9,]+", keep.digits)
range.pattern <- list(
  chrom="chr.*?",
  ":",
  chromStart=pos.pattern,
  list(
    "-",
    chromEnd=pos.pattern
  ), "?")
nc::capture_first_vec(pos.vec, range.pattern)
```

In summary, `nc::capture_first_vec` takes a variable number of arguments:

* The first argument is the subject character vector.
* The other arguments specify the pattern, via character strings,
  functions, and/or lists.
* If a pattern (character/list) is named, a capture group is generated in the regex.
* Each function is used to convert the text extracted by the previous
  named pattern argument.
* Lists may be used to avoid repetition in the definition of the pattern and type conversion functions.
* Each list generates a group in the regex (named list => named capture group, un-named list => non-capturing group).
* All patterns are pasted together in the order that they appear in
  the argument list.
  
## View generated regex

To see the generated regular expression pattern string, call
`nc::var_args_list` with the variable number of arguments that
specify the pattern:

```{r}
nc::var_args_list(range.pattern)
```

The generated regex is the `pattern` element of the resulting list
above. The other element `fun.list` indicates the names and type
conversion functions to use with the capture groups.

## Error/NA if any subjects do not match

The default is to stop with an error if any subject does not match:

```{r, error=TRUE, purl=FALSE}
bad.vec <- c(bad="does not match", pos.vec)
nc::capture_first_vec(bad.vec, range.pattern)
```

Sometimes you want to instead report a row of NA when a subject does
not match. In that case, use `nomatch.error=FALSE`:

```{r}
nc::capture_first_vec(bad.vec, range.pattern, nomatch.error=FALSE)
```

## Other regex engines

By default nc uses the PCRE regex engine. Other choices include ICU
and RE2. Each engine has different features, which are discussed in
[my R journal paper](https://github.com/tdhock/namedCapture-article).

The engine is configurable via the `engine` argument or the
`nc.engine` option:

```{r, error=TRUE, purl=FALSE}
u.subject <- "a\U0001F60E#"
u.pattern <- list(emoji="\\p{EMOJI_Presentation}")
old.opt <- options(nc.engine="ICU")
nc::capture_first_vec(u.subject, u.pattern)
nc::capture_first_vec(u.subject, u.pattern, engine="PCRE") 
nc::capture_first_vec(u.subject, u.pattern, engine="RE2")
options(old.opt)
```

## Capture first match in columns of a data.frame

We also provide `nc::capture_first_df` which extracts text
from several columns of a data.frame, using a different 
regular expression for each column.

* It requires a data.frame (or data.table) as the first argument.
* It takes a variable number of other arguments, all of which must be
  named. For each other argument we call `nc::capture_first_vec` on one
  column of the input data.frame.
* Each argument name specifies a column of the data.frame which will
  be used as the subject in `nc::capture_first_vec`.
* Each argument value specifies a pattern to be used with
  `nc::capture_first_vec`, in list/character/function format as
  explained in the previous section.
* The return value is a data.table with the same number of rows as the
  input, but with an additional column for each named capture
  group. New columns are named using the R argument names, so make
  sure they are unique.
* This function inputs data.frame and outputs data.table so can be used in a
  [pipe](https://r4ds.had.co.nz/pipes.html).
  
This function can greatly simplify the code required to create numeric
data columns from character data columns. For example consider the
following data which was output from the [sacct](https://slurm.schedmd.com/sacct.html) program.

```{r}
(sacct.df <- data.frame(
  Elapsed = c(
    "07:04:42", "07:04:42", "07:04:49",
    "00:00:00", "00:00:00"),
  JobID=c(
    "13937810_25",
    "13937810_25.batch",
    "13937810_25.extern",
    "14022192_[1-3]",
    "14022204_[4]"),
  stringsAsFactors=FALSE))
```

Say we want to filter by the total Elapsed time (which is reported as
hours:minutes:seconds), and base job id (which is the number before
the underscore in the JobID column). We could start by converting
those character columns to integers via:

```{r}
int.pattern <- list("[0-9]+", as.integer)
range.pattern <- list(
  "\\[",
  task1=int.pattern,
  list(
    "-",#begin optional end of range.
    taskN=int.pattern
  ), "?", #end is optional.
  "\\]")
nc::capture_first_df(sacct.df, JobID=range.pattern, nomatch.error=FALSE)
```
The result shown above is another data frame with an additional column
for each capture group. Next, we define another pattern that
matches either one task ID or the previously defined range pattern:

```{r}
task.pattern <- list(
  "_",
  list(
    task=int.pattern,
    "|",#either one task(above) or range(below)
    range.pattern))
nc::capture_first_df(sacct.df, JobID=task.pattern)
```

Below we match the complete JobID column:

```{r}
job.pattern <- list(
  job=int.pattern,
  task.pattern,
  list(
    "[.]",
    type=".*"
  ), "?")
nc::capture_first_df(sacct.df, JobID=job.pattern)
```

Below we match the Elapsed column with a different regex:

```{r}
elapsed.pattern <- list(
  hours=int.pattern,
  ":",
  minutes=int.pattern,
  ":",
  seconds=int.pattern)
nc::capture_first_df(sacct.df, JobID=job.pattern, Elapsed=elapsed.pattern)
```

Overall the result is another data table with an additional column for
each capture group. 
