<!--
%\VignetteEngine{knitr::knitr}
%\VignetteIndexEntry{vignette 3: capture melt}
-->

# Capture melt

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette explains how to use functions for "melting" wide
data tables, i.e. converting to tall/long data tables:

* `nc::capture_melt_single` is useful when you have a wide data table
  with lots of columns, all of the same type, that you would like to
  melt into a single result/output column.
* `nc::capture_melt_multiple` is useful when you have a wide
  data table with columns of different types that you would like to
  melt into multiple result/output columns.

Both are useful mainly in the case when you want to use a regular
expression to specify both (1) the set of input columns to reshape and
(2) some information to extract from those column names. To clarify
the discussion we introduce the following three terms:

* Reshape columns contain the data which are present in equal amounts
  in the input and output, but in fewer columns in the output.
  These are `measure.vars` in reshape2/data.table.
* Copy columns contain data which are copied from a single row of the
  input into several rows of the output. These are `id.vars`
  in reshape2/data.table.
* Capture columns are only present in the output, and contain data
  extracted from the names of the input reshape columns,
  using a capturing regular expression.
  As far as I know, `tidyr::pivot_longer` is the only other
  data reshaping function which directly supports
  capture columns.

## Capture data.frame column names and melt matching columns

Sometimes you want to melt a "wide" data table which has several
distinct pieces of information encoded in each column name. One
example is the familiar iris data, which have flower part and
measurement dimension encoded in each of four column names:

```{r}
library(data.table)
(iris.dt <- data.table(i=1:nrow(iris), iris[,1:4], Species=paste(iris$Species)))
```

Those four reshape column names can be specified via a regex in
`nc::capture_melt_single`. The first argument is the input data table
to reshape, and the subsequent arguments are interpreted as a pattern
which is passed to `nc::capture_first_vec`. Any input column names
which match the specified regex will be passed as `measure.vars` to
`melt`:

```{r}
(iris.tall <- nc::capture_melt_single(
  iris.dt,
  part=".*",
  "[.]",
  dim=".*"))
```

The output comes from joining the result of `nc::capture_first_vec` to
the result of `melt`. Note the output has two copy columns (i,
Species), one reshape column (value), and two capture columns (part,
dim).

After the data have
been converted to the tall/long table above, we can do a `dcast` to get a
wider data table:

```{r}
(iris.part.cols <- dcast(
  iris.tall,
  i + Species + dim ~ part))
```

Rather than use `capture_melt_single` and then `dcast`, we could
instead use `capture_melt_multiple` to get the same result:

```{r}
nc::capture_melt_multiple(
  iris.dt,
  column=".*",
  "[.]",
  dim=".*")
```

Note that the reshaped table above contains two copy columns (i,
Species), two reshape columns (Sepal, Petal), and one capture columns
(dim).  These can help us visualize whether or not sepals are bigger
than petals:

```{r, fig.width=10}

if(require(ggplot2)){
  ggplot()+
    theme_bw()+
    theme(panel.spacing=grid::unit(0, "lines"))+
    facet_grid(dim ~ Species)+
    coord_equal()+
    geom_abline(slope=1, intercept=0, color="grey")+
    geom_point(aes(
      Petal, Sepal),
      data=iris.part.cols)
}

```

It is clear from the plot above that sepals are indeed both longer and
wider than petals, on each measured plant.

Another data set where it is useful to do column name pattern matching
followed by melting is the World Health Organization data:

```{r}

if(requireNamespace("tidyr")){
  data(who, package="tidyr")
}else{
  who <- data.frame(id=1, new_sp_m5564=2, newrel_f65=3)
}
names(who)

```

Each column which starts with `new` has three distinct pieces of
information encoded in its name: diagnosis type (e.g. sp or rel),
gender (m or f), and age range (e.g. 5564 or 1524). We would like to
use a regex to match these column names, then using the matching
columns as measure.vars in a melt, then join the two results. 

```{r}
new.diag.gender <- list(
  "new_?",
  diagnosis=".*",
  "_",
  gender=".")
nc::capture_melt_single(who, new.diag.gender, ages=".*")
```

Note the output includes the new reshape column called `value` by
default, as in `melt`. The input reshape column
names which matched the specified pattern, and there is a new column
for each group in that pattern. The following example shows how to
rename the `value` column and use numeric type conversion functions:

```{r}
years.pattern <- list(new.diag.gender, ages=list(
  min.years="0|[0-9]{2}", as.numeric,
  max.years="[0-9]{0,2}", function(x)ifelse(x=="", Inf, as.numeric(x))))
(who.typed <- nc::capture_melt_single(
  who, years.pattern,
  value.name="count"))
str(who.typed)
```

The result above shows that 

* non-character captured output columns can be obtained by specifying
  type conversion functions in the pattern.
* `value.name` is used as the column name for the reshaped data in the
  output (as in `melt.data.table`).

In conclusion `nc::capture_melt_single` does data
reshaping from wide to tall/long which (1) makes it easy to define complex patterns (2)
supports type conversion without a post-processing step, and (3)
reduces repetition in user code.

## Melting columns of different types into multiple result columns

In the previous sections we discussed methods for melting wide data
columns that all have the same type. Sometimes wide data have columns
of different types that we want to melt. The example below is taken
from a data.table vignette,

```{r}
family.dt <- fread(text="
family_id age_mother dob_child1 dob_child2 dob_child3 gender_child1 gender_child2 gender_child3
1         30 1998-11-26 2000-01-29         NA             1             2            NA
2         27 1996-06-22         NA         NA             2            NA            NA
3         26 2002-07-11 2004-04-05 2007-09-02             2             2             1
4         32 2004-10-10 2009-08-27 2012-07-21             1             1             1
5         29 2000-12-05 2005-02-28         NA             2             1            NA")
```

The data table above has one row for each family, and two columns for
each child. One column is the date of birth (character), the other is
the gender (coded as an integer). We can use standard data.table
functions to melt these data into a tall/long table with one row for each
child:

```{r}
melt(family.dt, measure.vars=patterns(
  dob="^dob", gender="^gender"
))
```

The melt above results in a table with one row for each child, and one
column for each argument of `patterns`. However the code is somewhat
repetitive, as `dob` and `gender` must be repeated. To avoid this
repetition we can instead use:

```{r}
(children.dt <- nc::capture_melt_multiple(
  family.dt,
  column=".*",
  "_",
  nc::field("child", "", "[1-3]", as.integer),
  na.rm=TRUE))
str(children.dt)
```

The result above is similar to the previous result (and in fact
`melt.data.table` is used internally), but the code is less
repetitive. The first argument of `capture_melt_multiple` is the
subject data table and the following arguments form a pattern which is
matched to the input data column names. The pattern must have at least
two groups:

* One group must be named `column`. This group is used to capture the
  part of the input column names which will be used for the output
  column names. The output will contain one column for each unique
  value captured in this group.
* One or more other groups (with any names) specify an identifier for
  each observation (above it was the `child` number).

Another example is the iris data set which usually has one
5-dimensional observation per row (four numeric measures, and one
character `Species`). Below we create a shuffled version of the iris
data with two observations per row:

```{r}
set.seed(1)
iris.rand <- iris.dt[sample(.N)]
iris.wide <- cbind(treatment=iris.rand[1:75], control=iris.rand[76:150])
print(iris.wide, topn=2, nrows=10)
```

Below we show the usual data.table syntax for getting the original
iris back.

```{r}
iris.melted <- melt(iris.wide, measure.vars = patterns(
  i="i$",
  Sepal.Length="Sepal.Length$",
  Sepal.Width="Sepal.Width$",
  Petal.Length="Petal.Length$",
  Petal.Width="Petal.Width$",
  Species="Species$"))
identical(iris.melted[order(i), names(iris.dt), with=FALSE], iris.dt)
```

We can do the same thing with less repetition via:

```{r}
(nc.melted <- nc::capture_melt_multiple(
  iris.wide,
  group=".*?",
  "[.]",
  column=".*"))
identical(nc.melted[order(i), names(iris.dt), with=FALSE], iris.dt)
```

To conclude this section, we have seen that a simple non-repetitive
regex syntax can be used with `nc::capture_melt_multiple` to
melt wide data into several columns of different types.

